Hierarchical wrinkling in a confined permeable biogel

### 2015-01-20 fig_plis/article_plis.tex
Modifs Seb

\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{siunitx}
%\usepackage{multirow}

@@

\begin{document}
\title{Hierarchical wrinkling in a permeable and confined biogel}
\author{Mathieu Leocmach}
\affiliation{Universit\'e de Lyon, Laboratoire de Physique, \'Ecole Normale Sup\'erieure de Lyon, CNRS UMR 5672, 46 All\'ee d'Italie, 69364 Lyon cedex 07, France}
\author{Mathieu Nespoulous}
\affiliation{Aix-Marseille Université, CNRS, MADIREL UMR 7246, Marseille, France}
\author{Sébastien Manneville}
\affiliation{Universit\'e de Lyon, Laboratoire de Physique, \'Ecole Normale Sup\'erieure de Lyon, CNRS UMR 5672, 46 All\'ee d'Italie, 69364 Lyon cedex 07, France}
\author{Thomas Gibaud}
\email{thomas.gibaud@ens-lyon.fr}
\affiliation{Universit\'e de Lyon, Laboratoire de Physique, \'Ecole Normale Sup\'erieure de Lyon, CNRS UMR 5672, 46 All\'ee d'Italie, 69364 Lyon cedex 07, France}

\begin{abstract}

\end{abstract}

@@

\section*{Introduction}
There are many ways and reasons for a film to wrinkle. An elastic film buckles due to excess area with respect to its boundaries. Wrinkling is a buckling hindered by a substrate, the mismatch of elastic properties between the film and the substrate selects the preferred wrinkling wavelength, as discovered originally in aeronautics~\cite{Gough1940, Bijlaard1946}. On the engineering side, the last two decades have seen a bloom of methods to obtain well-controlled patterns via linear~\cite{Bowden1998, Genzer2006, Hu1998, Kim2010, Vandeparre2011, Li2013} or non-linear wrinkling~\cite{Efimenko2005, Guvendiren2010, Kim2011a, Brau2011}. Such patterns can be triggered by temperature dilation, swelling (eventually pH or light induced) or the removal of pre-strain. On the biology side, wrinkling-controlled morphogenesis is ubiquitous: ageing and the loss of elastic fibres makes our skin wrinkles~\cite{Bissett1987, Genzer2006}, difference in growth rates between the gut tube and its dorsal anchoring is responsible for the vilification of guts~\cite{Savin2011, Ciarletta2014, Shyer2013}, localized cell death in biofilms focuses mechanical forces and initiates 3D labyrinth pattern~\cite{Trejo2013, Asally2012}. On the physics side, analogy with wrinkling have been found in situations where an effective substrate comes from the resistance to uniaxial stretching~\cite{Cerda2003}, gravity~\cite{Kolinski2009, Vella2009, Pineirua2013, Lucantonio2013} (linked to the pioneering work of Smoluchowski~\cite{Smoluchowski1927}), boundaries~\cite{Vandeparre2011a, Li2013} or adhesion~\cite{Vella2009a}. Dealing only with (semi)infinite substrates, Biot generalised the elasticity of both the film and the substrate to viscoelasticity~\cite{Biot1957} or poroelasticity~\cite{Biot1964}, leading to time-dependent kinetic wavelength selection. Stability analysis of a film on a thin viscous substrate was studied more recently~\cite{Huang2002} and remains theoretical since on a free interface gravity dominates over viscous substrate.

Yet, in a biological context, thin tissues (epithelimum, endothelium) are immersed in a nearly buoyant medium (lymph, blood, mucus), a practical situation that could lead to wrinkling on a thin viscous substrate. Moreover, most biological films are porous, and little is known on the interplay between permeability and wrinkling~\cite{Ma2004,Longley2013}. Such interplay is also at stake in daily life problems such as wrinkles encountered while putting up wall paper~\cite{Pacione1991}.

Here, we have introduced permeable model bio-gels that produce hierarchical wrinkling as a result of the interplay between its gelation dynamics and the confinement conditions. In slit geometry, we slowly acidify a dispersion of casein, one of the milk proteins.  As a result, not only does the casein form a gel as expected~\cite{Leocmach2014} --this is a well know effect used in cheese or yogurt making, but, the gel starts to produce wrinkling patterns in cascade when casein adhesion to the slit wall is turned off. Using a combination of rheology, light microscopy and confocal microscopy, we demonstrate that, during the acidification, the casein network spontaneously shrink and swell at the micron level and wrinkles at the millimeter scale. We developed a model than pin down the dynamical origin of the constrains exerted on the gel and predict the wrinkling wavelength, Poiseuille~\cite{Poiseuille1842} flows of the solvent above and below the gel layer for long wavelengths and a new mechanism, Darcy~\cite{Darcy1856} flows of the solvent through the gel layer for short wavelengths.

\begin{figure}
	\includegraphics{dynamics}
	\caption{Dynamics of pattern formation. (a) Light transmission microscopy. The same area is highlighted in yellow on all panels to show that the initial pattern is not evolving. (b) 3D reconstruction from fluorescent confocal microscopy. (c) Fourier spectrum of binary altitudes of (a) corresponding to primary, secondary and tertiary patterns. Inset: reconstructed binary altitudes at \SI{2}{\hour}. Scale bars are \SI{1}{\milli\metre}.}
	\label{fig:dynamics}
\end{figure}

% § à déplacer après \section*{Pattern formation}, éventuellement en fusionnant des éléments avec le paragraphe précédent
In the present work, we study the swelling-induced wrinkling of a porous gel sheet sandwiched between two viscous layers, with the whole system confined between two glass plates. As shown in Fig.~\ref{fig:dynamics}, the lateral confinement leads to saturation of the amplitude on the floor and ceiling followed by a cascade of secondary, ternary, etc. buckling. Such cascade buckling have been observed by~\cite{Roman1999} in a one dimensional macroscopic system, a sheet of steel forced to buckle. The 2D nature of our experiment creates nested patterns that could prove useful to design optical elements. In addition, one has to note that the primary wavelength is much smaller than the size of the system, implying that we are not in a simple buckling scenario but in a wrinkling instability. Our main goal is to identify the selection mechanism of the wavelength.

% § à supprimer, ou à fusionner avec l'annonce de plan
In the following, we will describe the physical chemistry that allows us to form such patterns. Then, we will model the situation and identify two possible selection mechanisms: either Darcy flow or Poiseuille flow. Finally we will compare our experimental data with the two models. Combining both Darcy and Poiseuille mechanisms will allow us to describe our data.

@@

\begin{figure}[b]
	\includegraphics{acidification}
	\caption{Acid-induced protein gels properties behave non-monotonously with pH. (a) pH decrease in 4\%w sodium caseinate solution acidified by 4\%w GDL. Horizontal line indicates isoelectric pH of casen (b) Evolution of storage modulus measured in a rheometer (full adhesion to rotor and stator). (c) Evolution of pore size measured by confocal microscopy in a half coated microscopy cell (no adhesion to cell ceiling, syn\ae{}resis and swelling allowed).}
	\label{fig:acidification}
\end{figure}

@@

The slow acidification of a protein (sodium caseinate) solution down to the isoelectric pH induces gelation. Since caseins adsorb strongly to interfaces, the resulting gel sticks to container walls, allowing rheological measurements (see Fig.~\ref{fig:acidification}b). Further decrease of the pH beyond the isoeletric point weakens the gel. This non monotonous behaviour of the elastic modulus translates in a non monotonous mesh size in a microscopy cell (slit geometry) coated so that the casein could not adsorb on the top wall (Fig.~\ref{fig:acidification}c), i.e. in syn\ae{}resis (solvent expulsion) followed by swelling.

Figure~\ref{fig:sideview}b shows confocal measurements in the case where both top and bottom walls have been coated. Starting from an homogeneous solution (black margins correspond to the glass walls), the gel forms and simultaneously expels solvent on both sides. Since we have two clear interfaces between a protein rich phase and a protein poor (water) phase, we model the resulting situation as a porous elastic sheet of thickness $h$ sandwiched between two viscous layers of respective thicknesses $H_1$ et $H_2$ (see Fig.~\ref{fig:sideview}a). Relative increase in protein concentration during syn\ae{}resis can be deduced from the brighter fluorescence in second panel of Fig.~\ref{fig:sideview}b. Extracting the top and bottom surface of the gel phase allows for quantitative measurements of the relative volume (Fig.~\ref{fig:sideview}c) and excess surface area of the sheet of gel (Fig.~\ref{fig:sideview}d). We can thus confirm that the instability is due to swelling.

Note that adhesion to side walls is necessary to prevent total syn\ae{}resis and thus allow subsequent planar stresses while swelling. Also, the only situation were be observed wrinkling is when both top and bottom walls are coated. We confirmed that at any given time the thickness $h$ of the sheet of gel is homogeneous in $(x,y)$. The contrast in transmission microscopy (Fig.~\ref{fig:dynamics}a) is not due to thickness inhomogeneities but to altitude gradients. The length of the optical path (along $z$) through the gel phase is larger where the gel makes a slope compared to a flat situation, as indicated by the yellow vertical lines in Fig.~\ref{fig:sideview}a.

\begin{figure}
	\includegraphics{sideview}
	\caption{Model and measurements. (a) Schematic side view of the cell. The constant thickness gel sheet (orange) is surrounded by water. Yellow lines show the path of transmitted light through the gel phase. A region of interest is highlighted in gray. (b) Confocal XZ cuts showing syn\ae{}resis, swelling and (cascade) wrinkling. Scale bar is \SI{100}{\micro\metre} (real size ratio). Evolution of (c) volume of the gel phase relative to cell volume and (d) excess area measured by confocal microscopy.}
	\label{fig:sideview}
\end{figure}

\section*{Scalings}

The characteristic length $\lambda^*$ of a wrinkling pattern is set by the competition between the flexural modulus of the sheet $B=E/(1-\nu^2)$ ($E$ Young modulus, $\nu$ Poisson ratio) and a lateral load $\sigma_\perp$~\cite{Vella2009,Kolinski2009}: 
\begin{equation}
\lambda^* \sim h \left(\frac{B}{\sigma_\perp}\right)^{1/3}.
\label{eq:lstar}
\end{equation}
In small deflection actual wavelength $\lambda$ is related to the amplitude $A$ as $\lambda \sim A^{1/4} \lambda^{*3/4}$.

% Pour Mathieu : ajouter quelque part des applications numériques pour les trois longueurs d'ondes gravité, Darcy, Poiseuille
We identify three possible candidates for the lateral load $\sigma_\perp$: the weight or buoyancy of the gel, the porous flow through the gel or the viscous flows in the surrounding water layers. We confirmed that the former was playing no role by successfully repeating our experiment in a cell held vertical, i.e. with the weight acting longitudinally rather than laterally.

\begin{figure*}
	\includegraphics{Darcy_vs_Poiseuille}
	\caption{Comparing model predictions with measured wavelengths. Dots come from primary pattern, squares from secondary blisters. Dashed lines are the best linear fit through the origin taking only into account the points that should be (a) in Darcy mode $H<H^*$, (b) in Poiseuille mode $H>H^*$ (c) all points. Prefactors are 0.63, 0.69 and 0.67 respectively. Continuous line is the best affine fit ($\lambda_{xp}=0.52\lambda_{th}+\SI{0.33}{\milli\metre}$) to all the points in (b).}
	\label{fig:DarcyPoiseuille}
\end{figure*}

In a situation where the sheet of gel would be sitting (not sticking) to the floor of the cell ($H_1\rightarrow0$), creating a 2D blister requires flowing solvent of viscosity $\eta$ through the gel of permeability $\alpha$. Darcy law relates the gradient of pressure (identified to $\sigma_\perp/h$) to the volume flux per unit area $v = \frac{\alpha}{\eta}\frac{\sigma_\perp}{h} = \frac{dA}{dt}$. Injecting this expression in Eq.~(\ref{eq:lstar}) yields the wavelength of the Darcy mode as
\begin{equation}
\lambda_D \sim h^{1/2} \alpha^{1/4} \Upsilon^{1/4}
\end{equation}
where $\Upsilon = B\tau/\eta$ is a dimensionless viscoelastic factor quantifying the relative stiffness of the sheet and the solvent at a characteristic time $\tau$. This expression stresses the kinetic nature of the wavelength selection. However we note that the wavelength of (a generation of) the pattern does not evolve with time (see Fig.~\ref{fig:dynamics} and Sup. Fig.~\ref{fig:growth}) once the gel is touching both walls, we can deduce the relevant $\tau$ from the growth of the  pattern (see Supplementary Method for the precise definition).

The opposite limiting case is when the sheet of gel is in the middle of the cell ($H_1=H_2$) and impervious ($\alpha\rightarrow 0$), destabilisation over a wavelength $\lambda$ creates a lubrication ($H\ll\lambda$) flow in the viscous layers. By symmetry the lateral load across the gel $\sigma_\perp$ is also the pressure difference over the wavelength. Using a Poiseuille profile for the flow, $\frac{dA}{dt} \sim \sigma_\perp \frac{H^3}{\eta\lambda^2}$. Injecting this expression in Eq.~(\ref{eq:lstar}) yields the wavelength of the Poiseuille mode as
\begin{equation}
\lambda_P \sim (hH)^{1/2} \Upsilon^{1/6}
\end{equation}
where we recognise the scaling of the wrinkles of an elastic sheet on a thin ($H\ll\lambda$) elastic substrate~\cite{Cerda2003} where $\Upsilon$ would be the ratio between the flexural moduli of the film and substrate. Indeed, in lubrication or low Reynolds number one can consider a viscous film as elastic with effective Young modulus $3\eta/\tau$~\cite{Biot1957,Boudaoud2001}.

A more complete derivation of the models can be found in the Supplementary Method and yield the following prefactors:
\begin{align}
\lambda_D &= 2\pi h^{1/2}\alpha^{1/4}\left(\frac{\Upsilon}{12}\right)^{1/4}\\
\lambda_P &= \pi\sqrt{hH}\left(\frac{2}{9}\Upsilon\right)^{1/6}
\end{align}
where $1/H^3 = 1/H_1^3 + 1/H_2^3$. The limiting case between the two models is for $H = H^* \equiv 2^{2/3} 3^{1/6} \alpha^{1/2} \Upsilon^{1/6}$.

@@

The above models yield wavelengths of the same order in most of the cases we studied. Thus only systematic measurements of experimental $\lambda$ and all the parameters involved in either model allows to discriminate between them (see Supplementary Method for details on the measurements). As summarized in Supplementary Table~\ref{tab:data}, we changed the thickness of the cell, impacting $h$ and the maximal amplitude. We changed the composition and thus the stiffness of the gel and its permeability. We also changed systematically the viscosity of the solvent by adding glycerol, however this also impacted the gel stiffness, permeability and $\tau$. Finally, the only parameter we had no control on was the initial altitude of the gel before wrinklink ($H$) that to our knowledge varied randomly. However, we were able to obtain more data at small $H/H^*$ ratios by considering not only the primary patterns but also the secondary ones. Since the saturation of the primary pattern flattens the gel on the top and bottom walls, the secondary blisters always appear in a Darcy situation ($H\ll H^*$). Here the size of the blister is fixed as half of the size of the primary pattern as show on Fig.~\ref{fig:dynamics}c and $\tau$ is the variable parameter.

Our results are displayed in Fig.~\ref{fig:DarcyPoiseuille}. Neither model alone is able to reproduce all the data. However considering only the points where $H<H^*$ (respectively $H>H^*$), the Darcy (resp. Poiseuille) model fits the data well albeit with a prefactor. Note that it would be easy to oversight the failure of the Poiseuille model at small wavelength by adding an unexplained offset. Nonetheless, a complete model (derived in Supplementary Methods), summing up the influence of both Poiseuille and Darcy dissipation shows a better agreement for all values of $H/H^*$.

%Conclusion à écrire

>>>>

\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{siunitx}
%\usepackage{multirow}
\usepackage{bm}
\usepackage{dcolumn,ulem}
\usepackage{amsfonts}

@@

\newcommand{\seb}[1]{\textbf{\color{blue}#1}} % remarques Seb
\newcommand{\sseb}[1]{\sout{\textbf{\color{blue}#1}}} % remarques Seb

\begin{document}
\title{Hierarchical wrinkling in a confined permeable biogel}
\author{Mathieu Leocmach}
\affiliation{Universit\'e de Lyon, Laboratoire de Physique, \'Ecole Normale Sup\'erieure de Lyon, CNRS UMR 5672, 46 All\'ee d'Italie, 69364 Lyon cedex 07, France}
\author{Mathieu Nespoulous}
\affiliation{Aix-Marseille Université, CNRS, MADIREL UMR 7246, Marseille, France}
\author{Sébastien Manneville}
\affiliation{Universit\'e de Lyon, Laboratoire de Physique, \'Ecole Normale Sup\'erieure de Lyon, CNRS UMR 5672, 46 All\'ee d'Italie, 69364 Lyon cedex 07, France}
\author{Thomas Gibaud}
\email{thomas.gibaud@ens-lyon.fr}
\affiliation{Universit\'e de Lyon, Laboratoire de Physique, \'Ecole Normale Sup\'erieure de Lyon, CNRS UMR 5672, 46 All\'ee d'Italie, 69364 Lyon cedex 07, France}

\begin{abstract}
\seb{\it Ajouter l'abstract.}
\end{abstract}

@@

\section*{Introduction}

\seb{\it Terminologie generale a uniformiser dans tout l'article: sheet $\rightarrow$ layer, floor and ceiling $\rightarrow$ bottom and top walls, impervious $\rightarrow$ impermeable, non-monotonous $\rightarrow$ nonmonotonic, lateral load $\rightarrow$ transverse load, longitudinal/compression load $\rightarrow$ compressional load, laterally $\rightarrow$ transversely. A discuter puis a changer partout systematiquement ? De facon generale, il reste encore beaucoup de choses entre parentheses: soit c'est utile et on enleve les parentheses, soit c'est inutile et on supprime ?}

There are many ways and reasons for a film to wrinkle. An elastic film buckles due to excess area with respect to its boundaries. Wrinkling is a buckling hindered by a substrate: the mismatch of elastic properties between the film and the substrate selects the preferred wrinkling wavelength \sseb{as discovered originally in aeronautics} \cite{Gough1940, Bijlaard1946}. On the engineering side the last two decades have seen a bloom of methods to obtain well-controlled patterns via linear~\cite{Bowden1998, Genzer2006, Hu1998, Kim2010, Vandeparre2011, Li2013} or nonlinear wrinkling~\cite{Efimenko2005, Guvendiren2010, Kim2011a, Brau2011}. Such patterns can be triggered by temperature dilation, swelling (eventually pH or light induced) or the removal of pre-strain. On the biology side wrinkling-controlled morphogenesis is ubiquitous: ageing and the loss of elastic fibres makes our skin wrinkles~\cite{Bissett1987, Genzer2006}, difference in growth rates between the gut tube and its dorsal anchoring is responsible for the vilification of guts~\cite{Savin2011, Ciarletta2014, Shyer2013}, localized cell death in biofilms focuses mechanical forces and initiates 3D labyrinth pattern~\cite{Trejo2013, Asally2012}. On the physics side analogues of wrinkling have been found in situations where an effective substrate comes from the resistance to uniaxial stretching~\cite{Cerda2003}, from gravity~\cite{Kolinski2009, Vella2009, Pineirua2013, Lucantonio2013} \sseb{(linked to the pioneering work of Smoluchowski~\cite{Smoluchowski1927})} \seb{\it utile?}, from boundaries~\cite{Vandeparre2011a, Li2013} or from adhesion~\cite{Vella2009a}. Dealing only with (semi)infinite substrates, Biot generalised the elasticity of both the film and the substrate to viscoelasticity~\cite{Biot1957} or poroelasticity~\cite{Biot1964}, leading to time-dependent kinetic wavelength selection. Stability analysis of a film lying on a thin viscous substrate was studied only recently~\cite{Huang2002} \seb{and to date this confined situation remains without experimental equivalent.} \sseb{as for a free interface gravity dominates}
\sseb{ over viscous substrate} \seb{Pas tr\`es clair, \`a reformuler~?}

Yet, in a biological context, thin tissues (epithelimum, endothelium) are \seb{naturally confined and }immersed in a nearly buoyant medium (lymph, blood, mucus), a practical situation that could lead to wrinkling on a thin viscous substrate. Moreover, most biological films are porous, and little is known on the interplay between permeability and wrinkling~\cite{Ma2004,Longley2013}. \sseb{Such interplay is also at stake in daily life problems} 
\sseb{such as wrinkles encountered while putting up}
\sseb{ wall paper~\cite{Pacione1991}.} \seb{Therefore benchmark experiments that explore the possibility of wrinkling in confined soft materials immersed in a buoyancy-matched viscous medium are in line to get fundamental insights into such a biologically relevant situation.}

Here, we introduce model permeable biogels that produce hierarchical wrinkling as a result of their interplay between their gelation dynamics and the confinement conditions. In slit geometry, we slowly acidify a dispersion of casein, one of the milk proteins.  As a result, not only does the casein form a gel as expected~\cite{Leocmach2014} \seb{\it citer plutot une ref historique} --this is a well-known effect used in cheese or yogurt making-- but the gel starts to produce wrinkling patterns in cascade when casein adhesion to the slit wall is turned off. Using a combination of rheology, light microscopy and confocal microscopy, we demonstrate that upon acidification the casein network spontaneously shrinks then swells at the micron level and wrinkles at the millimeter scale. \seb{We systematically investigate such} swelling-induced wrinkling of a porous gel sheet sandwiched between two viscous layers, the whole system being confined between two glass plates. As shown in Fig.~\ref{fig:dynamics} \seb{and in Supplemental Movies}, lateral confinement leads to saturation of the amplitude on both the top and bottom walls followed by a cascade of secondary and ternary buckling. Such cascade buckling has already been observed by~\cite{Roman1999} in a one-dimensional macroscopic system, a sheet of steel forced to buckle \seb{between two solid plates}. The two-dimensional nature of our experiment creates nested patterns that could prove useful to design original optical elements. \seb{\it Mettre plutot la phrase precedente en discussion/conclusion ?} In addition one has to note that the primary wavelength is much smaller than the size of the system, implying that a simple buckling scenario \seb{cannot explain the observations in Fig.~\ref{fig:dynamics}} and that only a wrinkling instability \seb{can be invoked}.

\begin{figure}[ht!]
	\includegraphics{dynamics}
	\caption{Dynamics of pattern formation \seb{in a confined sheet of casein gel}. (a) Light transmission microscopy. The same area is highlighted in yellow on all panels to show that the initial pattern is not evolving. (b) 3D reconstruction from fluorescent confocal microscopy. (c) Fourier spectra of binary altitudes of (a) corresponding to primary, secondary and tertiary patterns. Inset: reconstructed binary altitudes at time \SI{2}{\hour}. Scale bars are \SI{1}{\milli\metre}. \seb{\it Indiquer la composition du gel~?}}
	\label{fig:dynamics}
\end{figure}

The main goal of the present study is thus to identify the selection mechanism of the wavelength  \seb{and to discuss its implication for practical situations (?)}. In the following we first describe the physical chemistry that allows us to form such patterns. Then we  model the situation and identify two possible selection mechanisms, namely porous Darcy flow or viscous Poiseuille flow. Comparing our experimental data to these two models we finally show that \sseb{a model based on} a combination of both Darcy and Poiseuille mechanisms accurately describes our data. This combined model pins down the dynamical origin of the constraints exerted on the gel and nicely predicts the wrinkling wavelength.
%Poiseuille~\cite{Poiseuille1842} flows of the solvent above and below the gel layer for long wavelengths and a new mechanism, Darcy~\cite{Darcy1856} flows of the solvent through the gel layer for short wavelengths.

@@

\begin{figure}[b]
	\includegraphics{acidification}
	\caption{Acid-induced protein gel properties behave nonmonotonically with pH. (a) pH decrease over time in a 4\%w sodium caseinate solution acidified by 4\%w GDL. Horizontal line \seb{The horizontal line? \it je laisse comme ca dans tous les captions.} indicates isoelectric pH of casein. (b) Evolution of the storage modulus $G'$ measured in a rheometer (full adhesion to rotor and stator). (c) Evolution of pore size measured by confocal microscopy in a half-coated microscopy cell (no adhesion to top wall, syn\ae{}resis and swelling allowed).}
	\label{fig:acidification}
\end{figure}

@@

The slow acidification of a protein (sodium caseinate) solution down to the isoelectric pH induces gelation (Fig.~\ref{fig:acidification}a). Since caseins strongly adsorb to interfaces, the resulting gel sticks to the container walls, allowing rheological measurements (Fig.~\ref{fig:acidification}b). Further decrease of the pH beyond the isoeletric point weakens the gel \seb{due to \dots} \seb{\it Il manque une explication ici ainsi que des references generales sur la prise du yaourt}. This nonmonotonic behaviour of the elastic modulus translates into a nonmonotonic mesh size in a microscopy cell (slit geometry) coated so that the casein cannot adsorb on the top wall (Fig.~\ref{fig:acidification}c), i.e. into syn\ae{}resis (solvent expulsion) followed by swelling.

Figure~\ref{fig:sideview}a shows confocal measurements in the case where both top and bottom walls have been coated. Starting from a homogeneous solution (black margins correspond to the glass walls), the gel forms and simultaneously expells solvent on both sides. Since two interfaces \seb{clearly separate} a protein-rich phase and a protein-poor (water) phase, we model the resulting situation as a porous elastic sheet of thickness $h$ sandwiched between two viscous layers of respective thicknesses $H_1$ and $H_2$ \seb{in a cell of thickness $e=H_1+h+H_2$} (Fig.~\ref{fig:sideview}d) \seb{\it $e$ n'etait defini nulle part + attention sur la Fig.~3c il vaudrait mieux choisir de montrer $H_1$ et $H_2$ a un endroit ou le gel est horizontal, sinon le schema n'est valable qu'aux petites deformations + du coup ce sera it mieux de mettre le schema en 3b ?}. Relative increase in protein concentration during syn\ae{}resis can be deduced from the brighter fluorescence in the second panel of Fig.~\ref{fig:sideview}a. Extracting the top and bottom surface of the gel phase allows for quantitative measurements of the relative volume (Fig.~\ref{fig:sideview}b) and excess surface area of the gel layer (Fig.~\ref{fig:sideview}c). This confirms that the instability is due to swelling.

Note that adhesion to side walls is necessary to prevent total syn\ae{}resis and thus allow subsequent planar stresses to develop while swelling: the only situation were wrinkling was observed is when both top and bottom walls are coated. We also confirmed that at any given \seb{point in time and space} the thickness $h$ of the sheet of gel is homogeneous. The contrast in transmission microscopy (Fig.~\ref{fig:dynamics}a) is \seb{therefore} not due to thickness inhomogeneities but to altitude gradients. \seb{Indeed} the length of the optical path (along \seb{the vertical direction} $z$) through the gel phase is larger where the gel makes a slope compared to a flat situation, as indicated by the yellow vertical lines in Fig.~\ref{fig:sideview}d.

\begin{figure}
	\includegraphics{sideview}
	\caption{Measurements and modeling. (a) Confocal ($x$,$z$) cuts showing syn\ae{}resis, swelling and (cascade) wrinkling. Scale bar is \SI{100}{\micro\metre} (real size ratio). Evolution of (b) volume of the gel phase relative to cell volume and (c) excess area measured by confocal microscopy. Crosses correspond to times in (a). (d) Schematic side view of the cell. The constant thickness gel layer (orange) is surrounded by water. Yellow lines show the path of transmitted light through the gel phase. A region of interest is highlighted in gray.}
	\label{fig:sideview}
\end{figure}

\section*{Models and scaling laws}

\seb{\it Ajouter une transition: en l'etat le passage au modele est trop brutal.}

The characteristic length $\lambda^*$ of a wrinkling pattern is set by the competition between the flexural modulus of the sheet $B=E/(1-\nu^2)$ ($E$ Young modulus, $\nu$ Poisson ratio) and a lateral load $\sigma_\perp$~\cite{Vella2009,Kolinski2009}: 
\begin{equation}
\lambda^* \sim h \left(\frac{B}{\sigma_\perp}\right)^{1/3}.
\label{eq:lstar}
\end{equation}
For small deflections the actual wavelength $\lambda$ is related to $\lambda^*$ and to the \seb{wrinkle} amplitude $A$ as $\lambda \sim A^{1/4} \lambda^{*3/4}$.

% Pour Mathieu : ajouter quelque part des applications numériques pour les trois longueurs d'ondes gravité, Darcy, Poiseuille
We identify three possible candidates for the lateral load $\sigma_\perp$: the weight or buoyancy of the gel, the porous flow through the gel or the viscous flows in the surrounding water layers. We confirmed that the former plays no significant role by successfully repeating our experiment in a cell held vertically, i.e. with the weight acting longitudinally rather than laterally.

\begin{figure*}
	\includegraphics{Darcy_vs_Poiseuille}
	\caption{Comparing model predictions with measured wavelengths. Dots come from primary pattern, squares from secondary blisters. Dashed lines are the best linear fit through the origin taking only into account the points that should be (a) in Darcy mode $H<H^*$, (b) in Poiseuille mode $H>H^*$ (c) all points. Prefactors are 0.63, 0.69 and 0.67 respectively. Continuous line is the best affine fit ($\lambda_{\rm exp}=0.52\lambda_{P}+\SI{0.33}{\milli\metre}$) to all the points in (b). \seb{\it Mettre $\lambda_P$ et $\lambda_D$ comme dans le texte + $\lambda_{\rm mixed}$ ? Mettre tous les fits de la meme couleur (en gris ou en noir). Ajouter les textes ``porous model'', ``viscous model'' et ``mixed model'' (ou equivalent) sur les graphes.}}
	\label{fig:DarcyPoiseuille}
\end{figure*}

In a first limit where the sheet of gel would be sitting (\seb{without} sticking) to the floor of the cell ($H_1\rightarrow0$), creating a 2D blister requires flowing solvent of viscosity $\eta$ through the gel of permeability $\alpha$. Darcy law relates the pressure gradient (\seb{estimated as} $\sigma_\perp/h$) to the volume flux per unit area $v = \frac{\alpha}{\eta}\frac{\sigma_\perp}{h} = \frac{dA}{dt}$. Injecting this expression in Eq.~(\ref{eq:lstar}) yields the wavelength of the Darcy mode as
\begin{equation}
\lambda_D \sim h^{1/2} \alpha^{1/4} \Upsilon^{1/4}
\end{equation}
where $\Upsilon = B\tau/\eta$ is a dimensionless viscoelastic factor quantifying the relative stiffness of the sheet and the solvent at a characteristic time $\tau$. This expression emphasizes the kinetic nature of the wavelength selection \seb{through Darcy flow}. However we note that the wavelength of (a generation of) the pattern does not evolve with time (see Fig.~\ref{fig:dynamics}a and Sup. Fig.~\ref{fig:growth}) once the gel is touching both walls, we can deduce the relevant $\tau$ from the growth of the  pattern (see Supplementary Method for the precise definition). \seb{\it La phrase precedente n'est pas assez claire et semble contredire le texte en supplementary : a reformuler.}

The opposite limiting case is when the gel layer lies in the middle of the cell ($H_1=H_2$) and impermeable ($\alpha\rightarrow 0$). There destabilisation over a wavelength $\lambda$ creates a lubrication ($H\ll\lambda$) flow in the viscous layers. By symmetry the lateral load across the gel $\sigma_\perp$ is also the pressure difference over the wavelength. Using a Poiseuille profile for the flow, $\frac{dA}{dt} \sim \sigma_\perp \frac{H^3}{\eta\lambda^2}$. Injecting this expression in Eq.~(\ref{eq:lstar}) yields the wavelength of the Poiseuille mode as
\begin{equation}
\lambda_P \sim (hH)^{1/2} \Upsilon^{1/6}
\end{equation}
where we recognize the scaling for the wrinkles of an elastic sheet on a thin ($H\ll\lambda$) elastic substrate~\cite{Cerda2003} where $\Upsilon$ would be the ratio between the flexural modulus of the film and that of the substrate. Indeed within the lubrication approximation or at low Reynolds numbers one can consider a viscous film as elastic with effective Young modulus $3\eta/\tau$~\cite{Biot1957,Boudaoud2001}.

A more complete derivation of the models can be found in the Supplementary Method and yields the following prefactors:
\begin{align}
\lambda_D &= 2\pi h^{1/2}\alpha^{1/4}\left(\frac{\Upsilon}{12}\right)^{1/4}\\
\lambda_P &= \pi (hH)^{1/2}\left(\frac{2}{9}\Upsilon\right)^{1/6}
\end{align}
where $1/H^3 = 1/H_1^3 + 1/H_2^3$. The crossover between the two models where $\lambda_D=\lambda_P$ is found for $H = H^* \equiv 2^{2/3} 3^{1/6} \alpha^{1/2} \Upsilon^{1/6}$.

\seb{\it Resumer ce qui precede en une phrase disant dans quel domaine de longueur d'onde on s'attend a trouver quel scaling pour nos manips.}

@@

The two above models yield wavelengths that are of the same order in most of the cases under study. Thus only systematic measurements of the experimental $\lambda$ \seb{together with estimates for} all the parameters involved in either model shall allow one to discriminate between them (see Supplementary Method for details on such measurements). As summarized in Supplementary Table~\ref{tab:data} we systematically varied (i) the thickness of the cell, impacting $h$ and the maximal amplitude, (ii) the gel composition and thus both its stiffness and its permeability, (iii) the viscosity of the solvent by adding glycerol, which also impacted the gel stiffness, permeability and $\tau$. Finally, the only parameter that we had no control upon was the initial altitude $H$ of the gel before wrinkling that varied randomly. However we were able to obtain more data at small $H/H^*$ ratios by considering not only the primary patterns but also the secondary ones. Since the saturation of the primary pattern flattens the gel on the top and bottom walls, the secondary blisters always appear in a Darcy situation ($H\ll H^*$). Here \seb{the wavelength associated to the secondary} blister is taken as half of that of the primary pattern as shown in Fig.~\ref{fig:dynamics}c and $\tau$ is the variable parameter. \seb{\it Pas sur que la fin de cette phrase soit bien claire pour moi.}

Our results are displayed in Fig.~\ref{fig:DarcyPoiseuille}. \seb{Although they provide the correct order of magnitude for $\lambda$,} neither model alone is able to \seb{quantitatively} reproduce the full data set. However considering only the points where $H<H^*$ (respectively $H>H^*$), the Darcy (resp. Poiseuille) model fits the data well albeit with a prefactor. Note that it would be easy to oversight \seb{\it ``oversight'' $\neq$ verbe: plutot ``oversee'' ?} the failure of the Poiseuille model at small wavelengths by adding an unexplained offset \seb{(dashed line in Fig.~\ref{fig:DarcyPoiseuille}b)}. Nonetheless, a complete model (derived in Supplementary Methods), summing up the influence of both Poiseuille and Darcy dissipation shows a better agreement for all values of $H/H^*$.

\seb{\it Ajouter une conclusion.}
